---
title: Fig5
subtitle:  "Number of chromatin accessibility peak changes during each transition (opening peaks, red; closing peaks, blue; FDR<0.05)."
author: "pz"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    highlight: tango
    df_print: paged
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',fig.width=10,tidy=TRUE, cache=FALSE)
knitr::opts_knit$set(progress=TRUE,verbose=TRUE)
```

```{r libraries, cache=FALSE}
library(data.table)
library(magrittr)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size = 16))
library(DESeq2)
library(tidyverse)
```



```{r decoder}
decoderFile <- "atac_decoder.csv"
decoder.data <- read.table(decoderFile,header=T,stringsAsFactors=F, sep=",")
decoder.data$SampleID = make.names(decoder.data$SampleID)
decoder.data$Condition = factor(decoder.data$Condition)
decoder.data$Condition <- relevel(decoder.data$Condition, ref = "N")
```


```{r loadDESeq2Object}
# see runDiffBind.R, runSVA.R
load("sva_dds_vsd.RData")
```

# Samples

The following samples were part of this analysis:

```{r samples}
as_tibble(decoder.data)
```

# Differential accessibility analysis


```{r da}
varInt="Condition"
dbs.all <- list()
alphaTR = 0.05
lfcThres = 0

comps = c("N_vs_E6hr", "E6hr_vs_E12hr", "E12hr_vs_E24hr", "N_vs_L6hr", "L6hr_vs_L12hr", "L12hr_vs_L1d")
for (comp in comps){
  levelRef <- strsplit(comp, "_vs_")[[1]][1]
  levelTest <- strsplit(comp, "_vs_")[[1]][2]
  dbs.all[[paste0(levelTest,"_vs_",levelRef)]] <- 
    results(dds, contrast=c(varInt, levelTest, levelRef), alpha=alphaTR) %>% as.data.frame()
}

names(dbs.all) = gsub("hr", "", names(dbs.all))
```


The following genes were detected as differentially expressed (adjusted p < `r alphaTR` and abs(log2FoldChange) > `r lfcThres`):

```{r printDEGtable}
dbs.all.sig <- lapply(names(dbs.all), function(x){
  subset(dbs.all[[x]], padj < alphaTR & abs(log2FoldChange) > lfcThres)
   })
names(dbs.all.sig) <- names(dbs.all)

res = data.frame(sapply(dbs.all.sig, NROW))
colnames(res) <-  paste0("padj<",alphaTR, " & ","abs(log2FoldChange)>",lfcThres)

rel.contrs = 1:nrow(res)

as_tibble(cbind(res,"contrast #"=rel.contrs) %>%  tibble::rownames_to_column(var="contrast"))
```

# Bar plot {.tabset}

Number of chromatin peak accessibility changes during each transition (FDR<0.05)

```{r fig.width=3, fig.height=5}
da_changes <- dbs.all.sig %>%
  imap(~{
    name <- paste(rev(strsplit(.y, "_vs_")[[1]]), collapse = "->")
    summarize(.x, contrast = name, positive = sum(log2FoldChange > 0), negative = sum(log2FoldChange < 0))
  }) %>%
  bind_rows() %>%
  pivot_longer(cols = c("positive", "negative"), names_to = "sign", values_to = "value") %>%
  mutate(contrast = factor(contrast, levels = c("N->E6", "E6->E12", "E12->E24", "N->L6", "L6->L12", "L12->L1d")), type = ifelse(grepl("L", contrast), "lesion", "effector"))
```

## Tumor

```{r fig.width=3, fig.height=5}
lesions <- da_changes %>% 
  filter(type == "lesion") %>% 
  mutate(contrast = factor(contrast, levels = c("N->L6", "L6->L12", "L12->L1d")))

as_tibble(lesions)

ggplot(lesions, aes(x = contrast)) + 
 geom_bar(data = subset(lesions, sign == "positive"),  width=0.45, aes(y = value/1000, fill = sign), stat = "identity", position = "dodge") +
geom_bar(data = subset(lesions, sign == "negative"),  width=0.45, aes(y = -value/1000, fill = sign), stat = "identity",  position = "dodge")+ 
  geom_hline(yintercept = 0,colour = "black") +
  scale_fill_manual(labels = c("Closing", "Opening"), values = c("#0019f1", "#e53222")) +
  labs(x = "", y = "# of peak changes (x 1k)") +
  theme_classic(base_size = 18) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.title = element_blank(),
        legend.key.size = unit(0.4, 'cm'),
        legend.position = c(0.80, 0.85),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank()) 
```

## Infection

```{r fig.width=3, fig.height=5}
effector <- da_changes %>% 
  filter(type == "effector") %>% 
  mutate(contrast = factor(contrast, levels = c("N->E6","E6->E12", "E12->E24")))

as_tibble(effector)

ggplot(effector, aes(x = contrast)) + 
 geom_bar(data = subset(effector, sign == "positive"),  width=0.45, aes(y = value/1000, fill = sign), stat = "identity", position = "dodge") +
geom_bar(data = subset(effector, sign == "negative"),  width=0.45, aes(y = -value/1000, fill = sign), stat = "identity",  position = "dodge")+ 
  geom_hline(yintercept = 0,colour = "black") +
  scale_fill_manual(labels = c("Closing", "Opening"), values = c("#0019f1", "#e53222")) +
  labs(x = "", y = "# of peak changes (x 1k)") +
  theme_classic(base_size = 18) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.title = element_blank(),
        legend.key.size = unit(0.4, 'cm'),
        legend.position = c(0.80, 0.85),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank()) 
```


# Session Info
```{r session, message=FALSE, warning=FALSE, cache=FALSE,echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```